---
output: html_document
title: "The Importance of Handwashing"
---

Dr. Ignaz Semmelweis is a Hungarian physician born in 1818 and active at the Vienna General Hospital. If Dr. Semmelweis looks troubled it's probably because he's thinking about childbed fever: A deadly disease affecting women that just have given birth. He is thinking about it because in the early 1840s at the Vienna General Hospital as many as 10% of the women giving birth die from it. He is thinking about it because he knows the cause of childbed fever: It's the contaminated hands of the doctors delivering the babies. And they won't listen to him and wash their hands!

In this notebook, we're going to reanalyze the data that made Semmelweis discover the importance of handwashing. Let's start by looking at the data that made Semmelweis realize that something was wrong with the procedures at Vienna General Hospital.

```{r include=FALSE, message=FALSE}

# Load packages

library(tidyverse)
library(kableExtra)

```
## Yearly Data
```{r echo=FALSE, message=FALSE}
# Read dataset

yearly <- read_csv("datasets/yearly_deaths_by_clinic.csv")

#view
yearly %>% 
    kable() %>% 
    kable_styling()
```


## The alarming number of deaths
<p>The table above shows the number of women giving birth at the two clinics at the Vienna General Hospital for the years 1841 to 1846. You'll notice that giving birth was very dangerous; an <em>alarming</em> number of women died as the result of childbirth, most of them from childbed fever.</p>
<p>We see this more clearly if we look at the <em>proportion of deaths</em> out of the number of women giving birth. </p>

```{r echo=FALSE, message=FALSE}
# Add a new column to yearly with proportion of deaths per no. births
yearly <- mutate(yearly, proportion_deaths = deaths/births)

# Print out yearly
yearly %>% 
    kable() %>% 
    kable_styling()
```
# Plot yearly proportion of deaths at the two clinics

```{r echo=FALSE}
yearly %>% 
ggplot(aes(x = year, y = proportion_deaths, colour = clinic))+
geom_line()

```

## Monthly Data

```{r echo=FALSE, message=FALSE}
# Read datasets/monthly_deaths.csv into monthly
monthly <- read_csv("datasets/monthly_deaths.csv")
head(monthly) %>% 
    kable() %>% 
    kable_styling()

# Adding a new column with proportion of deaths per no. births
monthly <- mutate(monthly, proportion_deaths = deaths/births)


# Print out the first rows in monthly
head(monthly) %>% 
    kable() %>% 
    kable_styling()

```


## Plot monthly proportion of deaths
```{r echo=FALSE}
monthly %>% 
ggplot(aes(x = date, y = proportion_deaths))+
geom_line()+
xlab ("Date")+
ylab ("Deaths per Birth")
```

## Mandatory Handwashing

```{r echo=FALSE, message=FALSE}
# From this date handwashing was made mandatory
handwashing_start = as.Date('1847-06-01')

# Add a TRUE/FALSE column to monthly called handwashing_started
monthly  <- monthly %>% 
mutate(handwashing_started = date >= handwashing_start)
head(monthly) %>% 
    kable() %>% 
    kable_styling()
```

## Plot of when Handwashing Started
```{r echo=FALSE}
# Plot monthly proportion of deaths before and after handwashing
monthly %>% 
ggplot(aes(x = date, y = proportion_deaths, color = handwashing_started))+
geom_line()+
xlab ("Date")+
ylab ("Deaths per Birth")

```

## Statistics
```{r echo=FALSE, message=FALSE}
# Calculating the mean proportion of deaths before and after handwashing.

monthly_summary <- monthly %>% 
group_by(handwashing_started) %>% 
summarise(mean(proportion_deaths))

# Printing out the summary.
monthly_summary %>% 
    kable() %>% 
    kable_styling()

# Calculating a 95% Confidence intrerval using t.test 
test_result <- t.test( proportion_deaths ~ handwashing_started, data = monthly)
test_result

```







